CodeBook
========================================================

The purpose of this project is to demonstrate your ability to collect, work with, and clean a data set. The goal is to prepare tidy data that can be used for later analysis. You will be graded by your peers on a series of yes/no questions related to the project. You will be required to submit: 1) a tidy data set as described below, 2) a link to a Github repository with your script for performing the analysis, and 3) a code book that describes the variables, the data, and any transformations or work that you performed to clean up the data called CodeBook.md. You should also include a README.md in the repo with your scripts. This repo explains how all of the scripts work and how they are connected.

One of the most exciting areas in all of data science right now is wearable computing - see for example this article . Companies like Fitbit, Nike, and Jawbone Up are racing to develop the most advanced algorithms to attract new users. The data linked to from the course website represent data collected from the accelerometers from the Samsung Galaxy S smartphone. A full description is available at the site where the data was obtained:

http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones

Here are the data for the project:

https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip

You should create one R script called run_analysis.R that does the following.

- Merges the training and the test sets to create one data set.
- Extracts only the measurements on the mean and standard deviation for each measurement.
- Uses descriptive activity names to name the activities in the data set.
- Appropriately labels the data set with descriptive activity names.
- Creates a second, independent tidy data set with the average of each variable for each activity and each subject.

Steps to reproduce this project
===============================================
1. Change the parameter of the setwd function call to the working directory/folder (i.e., the folder where these the R script file is saved).
2. Run the R script run_analysis.r. 

Input data description
==============================================
**Human Activity Recognition Using Smartphones Dataset **

**Version 1.0 **

The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain. See 'features_info.txt' for more details. 

For each record it is provided:

- Triaxial acceleration from the accelerometer (total acceleration) and the estimated body acceleration.
- Triaxial Angular velocity from the gyroscope. 
- A 561-feature vector with time and frequency domain variables. 
- Its activity label. 
- An identifier of the subject who carried out the experiment.

The dataset includes the following files:

- 'README.txt'

- 'features_info.txt': Shows information about the variables used on the feature vector.

- 'features.txt': List of all features.

- 'activity_labels.txt': Links the class labels with their activity name.

- 'train/X_train.txt': Training set.

- 'train/y_train.txt': Training labels.

- 'test/X_test.txt': Test set.

- 'test/y_test.txt': Test labels.

The following files are available for the train and test data. Their descriptions are equivalent. 

- 'train/subject_train.txt': Each row identifies the subject who performed the activity for each window sample. Its range is from 1 to 30. 

- 'train/Inertial Signals/total_acc_x_train.txt': The acceleration signal from the smartphone accelerometer X axis in standard gravity units 'g'. Every row shows a 128 element vector. The same description applies for the 'total_acc_x_train.txt' and 'total_acc_z_train.txt' files for the Y and Z axis. 

- 'train/Inertial Signals/body_acc_x_train.txt': The body acceleration signal obtained by subtracting the gravity from the total acceleration. 

- 'train/Inertial Signals/body_gyro_x_train.txt': The angular velocity vector measured by the gyroscope for each window sample. The units are radians/second. 

Output Data
==============================================
Output is tidy dataset file tidydata.txt containing data set with the average of each variable for each activity and each subject.

```{r}
str(tidata)
```

List the key variables in the data table
```{r}
key(tidata)
```

Summary of variables
```{r}
summary(tidata)
```


run_analysis.R 
=====================================================

This script performs all required data transformations to get tidy data as output: 

- Merges the training and the test sets to create one data set.
- Extracts only the measurements on the mean and standard deviation for each measurement.
- Uses descriptive activity names to name the activities in the data set.
- Appropriately labels the data set with descriptive activity names.
- Creates a second, independent tidy data set with the average of each variable for each activity and each subject.

## Download the file. Put it in the GettingandCleaningData folder.

```
url <- "http://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
  filename <- "Dataset.zip"
  
  path <- getwd()
  
  if(!file.exists(path))
  {
    dir.create(path)
  }
  
  download.file(url, file.path(path,filename))
```

## Load packages

```{r}
packages <- c("data.table", "reshape2")
sapply(packages, require, character.only = TRUE, quietly = TRUE)
```
## Read files to datatables

```{r}
## Read files to datatables
setwd("~/datasciencecoursera/GettingAndCleaningData/UCI HAR Dataset")
path <- getwd()

dtSubjectTrain <- fread(file.path(path, "train", "subject_train.txt"))
dtSubjectTest <- fread(file.path(path, "test", "subject_test.txt"))

dtYTrain <- fread(file.path(path, "train", "Y_train.txt"))
dtYTest <- fread(file.path(path, "test", "Y_test.txt"))

fileToDataTable <- function(f) {
  df <- read.table(f)
  dt <- data.table(df)
}

dtXTrain <- fileToDataTable(file.path(path, "train", "X_train.txt"))
dtXTest <- fileToDataTable(file.path(path, "test", "X_test.txt"))

```

##  Merge the training and the test sets to create one data set.
```{r}

## 1.  Merges the training and the test sets to create one data set.

dtSubject <- rbind(dtSubjectTest, dtSubjectTrain)
setnames(dtSubject, "V1", "subject")

dtY <- rbind(dtYTest, dtYTrain)
setnames(dtY, "V1", "activityNum")

dtX <- rbind(dtXTrain, dtXTest)

dtSubY <- cbind(dtSubject, dtY)

dt <- cbind (dtSubY, dtX)

setkey(dt, subject, activityNum)

head(dt, n=3)
names(dt)
```

## 2. Extract only the mean and standard deviation
```{r}
## 2. Extract only the mean and standard deviation

dtFeatures <- fread(file.path(path, "features.txt"))
setnames(dtFeatures, names(dtFeatures), c("ID", "featureName"))

dtFeatures <- dtFeatures[grepl("mean\\(\\)|std\\(\\)", featureName)]

dtFeatures$featureID <- dtFeatures[, paste0("V", ID)]

##Subset these variables using variable names.
select <- c(key(dt), dtFeatures$featureID)
dt <- dt[  ,select, with = FALSE]

names(dt)
```

## 3. Uses descriptive activity names to name the activities in the data set

```{r}
dtActivity <- fread(file.path(path, "activity_labels.txt"))
setnames (dtActivity, names(dtActivity), c("activityNum", "activityName"))
head(dtActivity)
```

## 4. Appropriately labels the data set with descriptive variable names. 

```{r}
## Adding activityName to dt1
dt1 <- merge(dtActivity, dt, by = "activityNum", all.y = TRUE)

setkey(dt1, subject, activityNum, activityName)

head(dt1, n=3)

## Reshaping data table
dtmelt <- data.table(melt(dt1, key(dt1), variable.name = "featureID"))
head(dtmelt, n=3)

## Adding featureName to data.table
dt2 <- merge(dtFeatures[, list(featureID, featureName)], dtmelt,  by = "featureID", 
             all.x = TRUE)
setkey(dt2, featureID, featureName, subject, activityNum, activityName) 

names(dt2)

head(dt2, n=3)
```

## 5. Creates a second, independent tidy data set with the average of each variable  for each activity and each subject. 
```{r}
tidata <- dcast(dt2, subject + activityNum + activityName ~ featureName, mean)
head(tidata, n=3)
## Write tidata to file "tidydata.txt"
write.table(tidata, file = "tidydata.txt", sep = " ", row.names = TRUE, col.names = TRUE)
```

